Temporal ordering and registration of images in studies of developmental dynamics

### 2014-05-27 image_analysis_paper/pnas_draft2.tex
added little edits to stuff

When time is the main source of variability,
vector diffusion maps can temporally order data; but it can also uncover multiple sources of variability.
%
%Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by %time.
%
%However, this is not always the case.
%
To illustrate this, we have collected a data set consisting of $90$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $48$ of the embryos are wild type, and $42$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, the algorithmic approaches to these tasks have been pursued relatively independently of each other. 
%
In particular, temporal ordering of imaging data was done manually for relatively small datasets, with a significant amount of human supervision, and using registered images as a starting point \cite{...}.  
%
In parallel, ordering large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different species, such as expression levels of different genes \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}. 
%
Automated temporal ordering of such datasets was done by first projecting them on a low-dimensional subspace, which can provided by principal or independent components, and then using this projection to solve a traveling salesman problem or construct a minimum spanning tree, providing in this way ordering for multiple developmental snapshots. 
%
The presented approach is different because it handles registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
%
Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.

The presented approach was demonstrated by applying it to three different experimental datasets, collected specifically for this study. 
%
In the first case, correct orientation and temporal order of images was known a priori, providing a test case for the algorithm.  
%
Both of these features were recovered by our technique with high accuracy. For this particular dataset, a simpler approach would suffice. 
%
Because of the relative simplicity of developmental dynamics (small tissue deformation and monotonic increase in the amplitude of a localized signaling pattern), temporal order of images could have been recovered by projection of registered images on the first principal component. 
%
This approach was clearly not sufficient for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
%
However, vector diffusion maps could readily extract a highly nontrivial developmental trajectory. 
%
Finally, we applied our algorithm to a heterogeneous dataset, composed of images from embryos that follow distinct developmental paths. 
%
We found that low-dimensional projection of registered data reveals these distinct paths and can correctly identify embryos of different genotypes.

Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector}. 
%
Because of the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
%
Based on our work, we think that a much smaller dataset may be needed for successful reconstruction of developmental trajectories, providing the first step in solving a large number of systems level tasks in studies of embryogenesis. 
%
In one of these tasks, the functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability.  
%
Another important application is related to the joint analysis of datasets provided by different imaging modalities, such as live imaging data of tissue morphogenesis and snapshots from fixed embryos stained with probes for specific molecular components.
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we are expect that dimensionality reduction approaches discussed in this work will find numerous applications and motivate future algorithmic advances.

@@

%\begin{itemize}
%\item What are other people doing  (maybe critically)
%\item Movies  smoothness
%\item Discrete symmetries like flips
%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
%
%%\item size is an issue
%\item  small tilts is an issue
%\item   could one do three-d image processing ?  of the floating thing, without the device
%\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
%\end{itemize}

>>>>

When time is the main source of variability,
vector diffusion maps can temporally order data; but it can also uncover multiple sources of variability.
%
%Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by %time.
%
%However, this is not always the case.
%
To illustrate this, we have collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, the algorithmic approaches to these tasks have been pursued relatively independently of each other. 
%
In particular, temporal ordering of imaging data was done manually for relatively small datasets, with a significant amount of human supervision, and using registered images as a starting point \cite{...}.  
%
In parallel, ordering large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different species, such as expression levels of different genes \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}. 
%
Automated temporal ordering of such datasets was done by first projecting them on a low-dimensional subspace, which can provided by principal or independent components, and then using this projection to solve a traveling salesman problem or construct a minimum spanning tree, providing in this way ordering for multiple developmental snapshots. 
%
In addition, the task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The presented approach is different because it handles registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
%
Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications \cite{zhao2003face}, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.

The presented approach was demonstrated by applying it to three different experimental datasets, collected specifically for this study. 
%
In the first case, correct orientation and temporal order of images was known {\it a priori}, which allowed us to validate the results of the algorithm.
%
Both the orientation and temporal order were recovered with high accuracy. 
%
However, for this particular dataset, a simpler approach would suffice. 
%
Because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern), temporal order could have been recovered by projection of registered images on the first principal component. 
%
However, PCA would clearly not suffice for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
%
However, vector diffusion maps could readily extract a highly nontrivial developmental trajectory. 
%
Finally, we applied our algorithm to a heterogeneous dataset, composed of images from embryos that follow distinct developmental paths. 
%
We found that low-dimensional projection of registered data reveals these distinct paths and can correctly identify embryos of different genotypes.

Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{shkolnisky2012viewing, singer2012vector}. 
%
Due to the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
%
Based on our work, we think that a much smaller dataset may be needed for successful reconstruction of developmental trajectories, providing the first step in solving a large number of systems level tasks in studies of embryogenesis. 
%
In one of these tasks, the functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability.  
%
Another important application is related to the joint analysis of datasets provided by different imaging modalities, such as live imaging data of tissue morphogenesis and snapshots from fixed embryos stained with probes for specific molecular components.
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we are expect that dimensionality reduction approaches discussed in this work will find numerous applications and motivate future algorithmic advances.

@@

\begin{itemize}
\item What are other people doing  (maybe critically)
\item Movies  smoothness
\item Discrete symmetries like flips
\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)

\item size is an issue
\item  small tilts is an issue
\item   could one do three-d image processing ?  of the floating thing, without the device
\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
\end{itemize}

### 2014-05-27 image_analysis_paper/si_information.tex
updated mutant figures and some text for paepr
added little edits to stuff

\begin{figure}[H]
\includegraphics[height=9cm]{raw_data3}
\caption{Images of cellularizing {\it Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
\end{figure}

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and the median of the pairwise distances between data points for the third data set.

@@

If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
As rotation matrices are orthonormal, many of these will be zero, 
and the nonzero inner products give us the embedding coordinates.
%
More specifically, the $j^{th}$ embedding coordinate will be given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle, \: 0 < k \le d, (j-1)d < l \le jd
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
and we select values of $k$ and $l$ such that the coordinate $\psi_{k, l}$ has the largest variability. 
%

@@

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
\item Construct the matrix $H$ as in \eqref{eq:H_to_R}.
\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
\end{enumerate}

@@

\begin{enumerate}
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
\item If we only want to order the data, then sort the data by the value of the entries in the second eigenvector
\item For higher-dimensional embeddings (e.g., the mutants), we use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
\end{enumerate}

@@

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
%
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
%
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
%
\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
%
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
%
\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
%
\item Compute the top few eigenvectors of $S$.
%
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
%
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
%
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
%
\item To compute the first (non-trivial) embedding coordinate for the data, compute coordinates as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
%
\item Use the $(k, l)$ pairing which gives the inner products with largest variance as the next embedding coordinate.
%
\item For successive coordinates, use successive blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$, compute the coordinates in \eqref{eq:vdm_coord}, and select the $(k, l)$ pairing with largest variance)

@@

Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a movie of the average developmental dynamics. 
%
Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
%
We define the frame at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter. 
%
For the images and movie in Figure~4, we take $\sigma^2 = 20$.
%

@@

\begin{figure}[H]
\includegraphics[width=5cm]{data1_coord_var}
\includegraphics[width=5cm]{data2_coord_var}
\includegraphics[width=5cm]{data3_coord_var}
\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
\end{figure}

>>>>

\begin{figure}[H]
\includegraphics[height=9cm]{raw_data3b}
\caption{Images of cellularizing {\it Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. $48$ of the images are of wild type embryos, and $42$ of the images are of mutant embryos.}
\end{figure}

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and half of the median of the pairwise distances between data points for the third data set.

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,l}$ such that $0 < k \le d, jd +1 < l \le (j+1)d$,
and we select the $(k,l)$ pair such that the coordinate $\psi_{k, l}$ has the largest variability. 
%

@@

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
\end{enumerate}

@@

\begin{enumerate}
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
\end{enumerate}

@@

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
%
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
%
\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
%
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
%
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
%
\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
%
\item Compute the top few eigenvectors of $S$.
%
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
%
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
%
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
%
\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
%
Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
%
\item To compute subsequent coordinates, use lower blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$). Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.

@@

Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a movie of the average developmental dynamics. 
%
Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
%
We define the frame at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter. 
%
For the images and movie in Figure~4, we take $\sigma^2 = 20$.
%

@@

%\begin{figure}[H]
%\includegraphics[width=5cm]{data1_coord_var}
%\includegraphics[width=5cm]{data2_coord_var}
%\includegraphics[width=5cm]{data3_coord_var}
%\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
%\end{figure}
